Incremental tactics

### 2012-06-08 doc/papers/LFMTP2012/paper.tex
paper: rename case to caseb (clashes with llncs)
paper: switch style from ACM to LNCS

\documentclass[9pt]{sigplanconf}

\usepackage{ntheorem}
\usepackage{amsmath}
\usepackage{amssymb}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\usepackage{macros}
\usepackage{mathpartir}
\usepackage{url}

@@

\conferenceinfo{LFMTP '12}{September 9, 2012 -- Copenhagen, Denmark}
\copyrightyear{2012}
\copyrightdata{[to be supplied]}

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Incremental tactics}
%\subtitle{Subtitle Text, if any}

\authorinfo{Matthias Puech}
           {Department of Comp. Sci., Univ. of Bologna,\\
             PPS, Team $\pi r^2$ (Univ. Paris Diderot, CNRS, INRIA)}
           {puech@cs.unibo.it}

\authorinfo{Yann R\'egis-Gianas}
           {PPS, Team $\pi r^2$ (Univ. Paris Diderot, CNRS, INRIA)}
           {yrg@pps.jussieu.fr}

@@

%% Here is a problem
\begin{abstract}
 %% 4 sentences:
 %% State the problem
 %% Say why it’s an interesting problem
 %% Say what your solution achieves
 %% Say what follows from your solution
\end{abstract}

\category{D.3.3}{Language Constructs and Features}{Data types and
  structures} \category{F.3.1}{Logics and Meanings of
  Programs}{Specifying and Verifying and Reasoning about
  Programs --- Logics of programs}

\terms
Theory, Languages

\keywords
incrementality, type checking, logical framework, version control

@@

Several generic methods have been devised for turning a batch program
into an incremental one \cite{pugh1989incremental,
  acar2003selective}. Unfortunately, they all suffer from their
genericity on the particular problem of type checking:
\begin{itemize}
\item first, they ignore the higher-order nature of program syntax,
  and the kind of structural quotients we make on syntax trees with
  binders and environments ($\alpha$-equivalence, weakening\ldots): a
  hypothetical derivation of\ \ $\vdash x+2 : \cst{nat}$ is still valid
  when put under a binder\ \ $\vdash\lam y (x + 2) -
  y:\cst{nat}\to\cst{nat}$ and should not be regenerated;
\item for very large programs, we might want to feed the type checker
  only with a small \emph{delta} on the previous version and not the
  whole program, thus saving the time to recognize already checked
  parts;
\item finally, there is no way to \emph{a posteriori} validate the
  result of the incremental process: the user has to trust not only
  the type-checking algorithm but also its transformation into an
  incremental one, which increases the size of the trusted base.
\end{itemize}

@@

Seeing type-checking as issuing a completely explicit typing
derivation enables thus to think of the relationship between the input
and the output. Feeding subterms of the output $\md_i$ back into
subterms of the input $P_i$ is a well-known technique for handling
changes in the input: combined with a way to automatically recognize
already-computed input, it is \emph{memoization}, a general-purpose
technique for incremental computation that has been extensively
studied and combined with other techniques \cite{pugh1989incremental,
  acar2003selective}. However, it is best understood in the case of
purely first-order data, and to our best knowledge remains unexplored
for higher-order structures containing binders, like programs and
proofs.

@@

\begin{mathleft}
  \lamd M \match{M} \\
  \quad\case{\z} \pair {\cst{even}}{\infer{ }{\vdash \cst o : \cst{even}}} \\
  \quad\case{\s M}
  \match{\finfer M} \\
  \quad\quad\case{\pair{\cst{even}}\md}
  \pair {\cst{odd}}{\infer\md{\vdash\app{\cst{s}} M : \cst{odd}}} \\
  \quad\quad\case{\pair{\cst{odd}}\md}
  \pair {\cst{even}}{\infer\md{\vdash\app{\cst{s}} M : \cst{even}}} \\
  \quad\quad\case{\pair{\cst{nat}}\md}
  \pair {\cst{nat}}{\infer\md{\vdash\app{\cst{s}} M : \cst{nat}}} \\
\end{mathleft}

@@

\begin{mathleft}
  \quad\case{\app{M}N} \\
  \quad\quad\letd {\pair {A_1\to B} {\md_1}} {\finfer{M}} \\
  \quad\quad\letd {\pair {A_2} {\md_2}} {\finfer{N}} \\
  \quad\quad\letd {\md_{\leq}} {\fleq {A_1} {A_2}} \\
  \quad\quad
  \match{\md_{\leq}} \\
  \quad\quad\quad\case{\infer{ }{\vdash A\leq A}}
  \pair {B} {\infer{\md_1 \and \md_2}{\vdash \app M N : B}} \\
  \quad\quad\quad\case{\_}
  \pair {B} {\infer{\md_1 \and \infer*{\md_2 \and \md_{\leq}}{\vdash N
      : A_1}}{\vdash \app M N : B}}
\end{mathleft}

@@

\begin{mathleft}
  \quad\case{\tlam x A M} \\
  \quad\quad
  \letdin{\pair B \md}{\env {$\md_{\var x}$} {(\vdash {\var x} : A)}}
  {\finfer {\gsubst M {\msubst x {\fget{\pair A {\md_{\var x}}}}}}}
\end{mathleft}

@@

\begin{mathleft}
  \quad\case{\recb M N x y P} \\
  \quad\quad
  \letd {\pair {A_M} {\md_M}} {\finfer{M}} \\
  \quad\quad\letd {\md_{A_M}} {{A_M} \leq \cst{nat}} \\
  \quad\quad
  \letd {\pair {A_N} {\md_N}} {\finfer{N}} \\
  \quad\quad
  \syntax{let} {\pair {A_P} {\md_P}} \syntax{in}
  {(\envcons {\env {$\md_x$} {(\vdash
        \var x : \cst{nat})}} {$\md_y$} {(\vdash\var y : A_N)})} = \\
  \quad\quad\quad
  {\finfer{\gsubst P {\msubstcons {\msubst x
          {\fget{\pair{\cst{nat}}{\md_x}}}} y
        {\fget{\pair{A_N}{\md_y}}}}}} \syntax{in}\\
  \quad\quad\letd {\pair A {\pair{\md_{A_N}}{\md_{A_P}}}} {{A_N} \sqcap {A_P}} \\
  \quad\quad
  \syntax{let} {\pair {\_} {\md_P}} \syntax{in}
  {(\envcons {\env {$\md_x$} {(\vdash
        \var x : \cst{nat})}} {$\md_y$} {(\vdash\var y : A)})} = \\
  \quad\quad\quad
  {\finfer{\gsubst P {\msubstcons {\msubst x
          {\fget{\pair{\cst{nat}}{\md_x}}}} y
        {\fget{\pair{A}{\md_y}}}}}} \syntax{in}\\
  \hspace{-2em}\pair {A} {
    \infer{
      \infer*{\md_M \and \md_{A_M}}
      {\vdash M : A} \and
      \infer*{\md_N \and \md_{A_N}}
      {\vdash N : A}\and
      \infer*{\infer*{}{\mbox{$
        \begin{array}{c}
          [\md_x] [\md_y] \\
          \md_P
      \end{array}
        $}} \and \md_{A_P}}{\vdash P : A}
    }{
      \vdash \recb M N x y P : A
    }
  }
\end{mathleft}

@@

We now turn to the definition of our computation calculus. As we saw,
code is attached to special, \emph{defined} constants in the signature
that manipulates \LF\ objects. We have the ability to construct
objects out of other, and deconstruct them by case analysis:
\begin{align*}
  T &\gequal
  \lam x T \gor
  U \\
  U &\gequal
  F \gor
  \matchin U \Gamma C \\
  C &\gequal \casenobar P U \gor
  C\ \case P U \\
  P &\gequal
  \app H {\var x} \ldots\ {\var x}
\end{align*}

@@

\begin{align*}
  \gsubst {(\lam x T)} \sigma &= \lam x {\gsubst T\sigma} \qquad
  \text{if $\var x\notin\dom\sigma$}\\
  \gsubst {(\matchin U \Gamma C)} \sigma &=
  \matchin {\gsubst
    U\sigma} {\gsubst\Gamma\sigma} {\gsubst C\sigma} \\
  \gsubst {(\casenobar{\app H {\spinecons {\var x_1} {\spinecons \ldots {\var
          x_n}}}} U)} \sigma &= {\casenobar{\app H {\spinecons {\var x_1} {\spinecons \ldots {\var
          x_n}}}} \gsubst U \sigma} \quad \text{if $\var x_i\notin\dom\sigma$} \\
\end{align*}

@@

    \infer{
      \jcr\mr{\gconcat\Gamma{\Gamma'}} U \mmr {\app {H_i} {\spinecons
          {M_1}{\spinecons\ldots{M_{m_i}}}}}
      \and
      \jcr\mmr\Gamma {\gsubst{U_i}{\msubst {x$_1$} {M_1} \ldots\
          \msubst {x$_{m_i}$} {M_{m_i}}}} \mmmr F a
    }{
      \jcr\mr\Gamma{\matchin{U}{\Gamma'}{
          \casenobar{\app {H_1}{\spinecons {\var x_{11}} {\spinecons
                \ldots\ {\var x_{1m_1}}}}}{U_1}
          |\ \ldots\
          \case{\app {H_n}{\spinecons {\var x_{n1}} {\spinecons
              \ldots\ {\var x_{nm_n}}}}}{U_n}
        }}\mmr F
    }

@@

\bibliographystyle{abbrvnat}
\bibliography{../../english.bib}

>>>>

\documentclass{llncs}

\usepackage{ntheorem}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{macros}
\usepackage{mathpartir}
\usepackage{url}

\allowdisplaybreaks[1]

@@

\title{Incremental tactics}

\author{Matthias Puech\inst{1,2} \and Yann R\'egis-Gianas\inst{1}}
\institute{
  PPS, Team $\pi r^2$ (Universit\'e Paris Diderot, CNRS, INRIA)
  \and
  Department of Computer Science, University of Bologna
}

@@

\begin{abstract}
 %% 4 sentences:
 %% State the problem
 %% Say why it’s an interesting problem
 %% Say what your solution achieves
 %% Say what follows from your solution
\end{abstract}

@@

Several generic methods have been devised for turning a batch program
into an incremental one
\cite{pugh1989incremental,acar2003selective}. Unfortunately, they all
suffer from their genericity on the particular problem of type
checking:
\begin{itemize}
\item first, they ignore the higher-order nature of program syntax,
  and the kind of structural quotients we make on syntax trees with
  binders and environments ($\alpha$-equivalence, weakening\ldots): a
  hypothetical derivation of\ \ $\vdash x+2 : \cst{nat}$ is still valid
  when put under a binder\ \ $\vdash\lam y (x + 2) -
  y:\cst{nat}\to\cst{nat}$ and should not be regenerated;
\item for very large programs, we might want to feed the type checker
  only with a small \emph{delta} on the previous version and not the
  whole program, thus saving the time to recognize already checked
  parts;
\item finally, there is no way to a posteriori validate the result of
  the incremental process: the user has to trust not only the
  type-checking algorithm but also its transformation into an
  incremental one, which increases the size of the trusted base.
\end{itemize}

@@

Seeing type-checking as issuing a completely explicit typing
derivation enables thus to think of the relationship between the input
and the output. Feeding subterms of the output $\md_i$ back into
subterms of the input $P_i$ is a well-known technique for handling
changes in the input: combined with a way to automatically recognize
already-computed input, it is \emph{memoization}, a general-purpose
technique for incremental computation that has been extensively
studied and combined with other techniques
\cite{pugh1989incremental,acar2003selective}. However, it is best
understood in the case of purely first-order data, and to our best
knowledge remains unexplored for higher-order structures containing
binders, like programs and proofs.

@@

\begin{mathleft}
  \lamd M \match{M} \\
  \quad\caseb{\z} \pair {\cst{even}}{\infer{ }{\vdash \cst o : \cst{even}}} \\
  \quad\caseb{\s M}
  \match{\finfer M} \\
  \quad\quad\caseb{\pair{\cst{even}}\md}
  \pair {\cst{odd}}{\infer\md{\vdash\app{\cst{s}} M : \cst{odd}}} \\
  \quad\quad\caseb{\pair{\cst{odd}}\md}
  \pair {\cst{even}}{\infer\md{\vdash\app{\cst{s}} M : \cst{even}}} \\
  \quad\quad\caseb{\pair{\cst{nat}}\md}
  \pair {\cst{nat}}{\infer\md{\vdash\app{\cst{s}} M : \cst{nat}}} \\
\end{mathleft}

@@

\begin{mathleft}
  \quad\caseb{\app{M}N} \\
  \quad\quad\letd {\pair {A_1\to B} {\md_1}} {\finfer{M}} \\
  \quad\quad\letd {\pair {A_2} {\md_2}} {\finfer{N}} \\
  \quad\quad\letd {\md_{\leq}} {\fleq {A_1} {A_2}} \\
  \quad\quad
  \match{\md_{\leq}} \\
  \quad\quad\quad\caseb{\infer{ }{\vdash A\leq A}}
  \pair {B} {\infer{\md_1 \and \md_2}{\vdash \app M N : B}} \\
  \quad\quad\quad\caseb{\_}
  \pair {B} {\infer{\md_1 \and \infer*{\md_2 \and \md_{\leq}}{\vdash N
      : A_1}}{\vdash \app M N : B}}
\end{mathleft}

@@

\begin{mathleft}
  \quad\caseb{\tlam x A M} \\
  \quad\quad
  \letdin{\pair B \md}{\env {$\md_{\var x}$} {(\vdash {\var x} : A)}}
  {\finfer {\gsubst M {\msubst x {\fget{\pair A {\md_{\var x}}}}}}}
\end{mathleft}

@@

\begin{mathleft}
  \quad\caseb{\recb M N x y P} \\
  \quad\quad
  \letd {\pair {A_M} {\md_M}} {\finfer{M}} \\
  \quad\quad\letd {\md_{A_M}} {{A_M} \leq \cst{nat}} \\
  \quad\quad
  \letd {\pair {A_N} {\md_N}} {\finfer{N}} \\
  \quad\quad
  \syntax{let} {\pair {A_P} {\md_P}} \syntax{in}
  {(\envcons {\env {$\md_x$} {(\vdash
        \var x : \cst{nat})}} {$\md_y$} {(\vdash\var y : A_N)})} = \\
  \quad\quad\quad
  {\finfer{\gsubst P {\msubstcons {\msubst x
          {\fget{\pair{\cst{nat}}{\md_x}}}} y
        {\fget{\pair{A_N}{\md_y}}}}}} \syntax{in}\\
  \quad\quad\letd {\pair A {\pair{\md_{A_N}}{\md_{A_P}}}} {{A_N} \sqcap {A_P}} \\
  \quad\quad
  \syntax{let} {\pair {\_} {\md_P}} \syntax{in}
  {(\envcons {\env {$\md_x$} {(\vdash
        \var x : \cst{nat})}} {$\md_y$} {(\vdash\var y : A)})} = \\
  \quad\quad\quad
  {\finfer{\gsubst P {\msubstcons {\msubst x
          {\fget{\pair{\cst{nat}}{\md_x}}}} y
        {\fget{\pair{A}{\md_y}}}}}} \syntax{in}\\
  \hspace{-2em}\pair {A} {
    \infer{
      \infer*{\md_M \and \md_{A_M}}
      {\vdash M : A} \and
      \infer*{\md_N \and \md_{A_N}}
      {\vdash N : A}\and
      \infer*{\infer*{}{\mbox{$
        \begin{array}{c}
          [\md_x] [\md_y] \\
          \md_P
      \end{array}
        $}} \and \md_{A_P}}{\vdash P : A}
    }{
      \vdash \recb M N x y P : A
    }
  }
\end{mathleft}

@@

We now turn to the definition of our computation calculus. As we saw,
code is attached to special, \emph{defined} constants in the signature
that manipulates \LF\ objects. We have the ability to construct
objects out of other, and deconstruct them by case analysis:
\begin{align*}
  T &\gequal
  \lam x T \gor
  U \\
  U &\gequal
  F \gor
  \matchin U \Gamma C \\
  C &\gequal \casebnobar P U \gor
  C\ \caseb P U \\
  P &\gequal
  \app H {\var x} \ldots\ {\var x}
\end{align*}

@@

\begin{align*}
  \gsubst {(\lam x T)} \sigma &= \lam x {\gsubst T\sigma} \qquad
  \text{if $\var x\notin\dom\sigma$}\\
  \gsubst {(\matchin U \Gamma C)} \sigma &=
  \matchin {\gsubst
    U\sigma} {\gsubst\Gamma\sigma} {\gsubst C\sigma} \\
  \gsubst {(\casebnobar{\app H {\spinecons {\var x_1} {\spinecons \ldots {\var
          x_n}}}} U)} \sigma &= {\casebnobar{\app H {\spinecons {\var x_1} {\spinecons \ldots {\var
          x_n}}}} \gsubst U \sigma} \quad \text{if $\var x_i\notin\dom\sigma$} \\
\end{align*}

@@

    \infer{
      \jcr\mr{\gconcat\Gamma{\Gamma'}} U \mmr {\app {H_i} {\spinecons
          {M_1}{\spinecons\ldots{M_{m_i}}}}}
      \and
      \jcr\mmr\Gamma {\gsubst{U_i}{\msubst {x$_1$} {M_1} \ldots\
          \msubst {x$_{m_i}$} {M_{m_i}}}} \mmmr F a
    }{
      \jcr\mr\Gamma{\matchin{U}{\Gamma'}{
          \casebnobar{\app {H_1}{\spinecons {\var x_{11}} {\spinecons
                \ldots\ {\var x_{1m_1}}}}}{U_1}
          |\ \ldots\
          \caseb{\app {H_n}{\spinecons {\var x_{n1}} {\spinecons
              \ldots\ {\var x_{nm_n}}}}}{U_n}
        }}\mmr F
    }

@@

\bibliographystyle{splncs}
\bibliography{../../english}
